The truncated KdV framework for modeling anomalous waves induced by abrupt depth changes

### 2019-07-15 WavesData-2019/LatexNotes/notes.tex
NM

%----------------------------------------------------------%
% Mathematical framework
%----------------------------------------------------------%
\section{The truncated KdV framework}

\subsection{The Korteweg–de Vries equation with variable depth}
Consider waves propagating unidirectionally in shallow water. Consider the surface displacement $\eta(x,t)$ and the reference frame moving with the local wave speed $\xi = x - ct$, where $c = \sqrt{g \depth}$ is the wave speed, $g$ gravity, and $\depth$ the local depth.
To first-correction in small amplitude, surface displacements are governed by the Korteweg–de Vries equation (KdV), which in dimensional form is given by
\begin{equation}
2 \eta_t + \frac{3 c}{\depth} \eta \eta_{\xi} + \frac{c \depth^2}{3} \eta_{\xi \xi \xi} = 0
\end{equation}
% The coefficients and scales can be verified on Wolfram KdV page (but what about the sign?). I would like an official source as confirmation though.

@@

% Nondimensionalization
\subsection{Dimensionless variables}
\label{nondim}

% Table
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{table}[h]%[htbp]
\begin{center}
\caption{Table of parameters}
\label{paramtable}
\begin{tabular}{l l l}
\hline \multicolumn{3} { c }{Parameters that are constant in a single experiment} \\
\hline Description & Notation and definition & Value in experiments \\
\hline
Peak forcing frequency		& $f_p$						& 2 Hz \\
Upstream depth			& $\dup$						& 12.5 cm \\
Downstream depth			& $\ddn$						& 3 cm \\
Characteristic wave amplitude	& $a = \mean{\eta^2}^{1/2} $		& 0.03--0.3 cm \\
Upstream wavelength		& $\lamup = \sqrt{g \dup}/f_p$		& 56 cm \\
Amplitude-to-depth ratio		& $\epsup = a / \dup$			& 0.0024--0.024 \\
Depth-to-wavelength ratio		& $\delup = \dup / \lamup$		& 0.22 \\
Depth ratio				& $\dratdn = \ddn/\dup$			& 0.24 \\
\hline \multicolumn{3} { c }{Parameters that change value across ADC} \\
\hline Description & Notation and definition \\
\hline
Local depth			& $\depth$			\\
Local wavespeed		& $c = \sqrt{g \depth}$	\\
Local peak wavelength	& $\lam = \sqrt{gd}/f_p$	\\
Local dimensionless depth		& $\drat = \depth/\dup$	
\end{tabular}
\end{center}
\end{table}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

@@

\subsubsection{Alternative KdV formulation following Johnson}
Section \ref{nondim} makes the implicit assumption of continuous surface displacement crossing the ADC. This is a very physically motivated assumption, but there is a small amount of debate surrounding the assumption. An alternative is to appeal to Green's law and assume the quantity $\depth^{1/4} \eta$ to be continuous across the ADC, which necessarily implies a small discontinuity in the surface (which of course cannot be true literally), as is the approach taken in Johnson't book. The dimensional KdV equation would then be expressed as
\begin{equation}
2 (\depth^{1/4} \eta)_t + \frac{3 c}{\depth^{5/4}} (\depth^{1/4} \eta) (\depth^{1/4} \eta)_{\xi} + \frac{c \depth^2}{3} (\depth^{1/4} \eta)_{\xi \xi \xi} = 0
\end{equation}
Then, the dimensionless state variable is instead 
\begin{equation}
u = \frac{\depth^{1/4} \, \eta}{ \dup^{1/4} \, \amp}
\end{equation}
and the dimensionless KdV equation takes the form
\begin{equation}
u_t + \frac{3 \pi}{\lamfac} \epsup \drat^{-5/4} \, u u_x + \frac{4 \pi^3}{3 \lamfac^3} \delup^2 \drat \, u_{xxx} = 0
\qquad \text{for } x \in [-\pi,\pi]
\end{equation}
Notice the only difference is a small change in the power of $\drat$ in the second term (changes from $\drat^{-1}$ to $\drat^{-5/4}$).
We will not use this alternative formulation.

\subsection{Hamiltonian structure}

@@

We measure the state variable $u$ as the displacement of the surface from equilibrium, so that $\mean{u} = 0$. Consequently, the momentum of $u$ vanishes
\begin{equation}
M[u] \defeq \int_{-\pi}^{\pi} u \dx = 0
\end{equation}
Next, the definition of the characteristic amplitude in \eqref{ampdefn} implies that $\mean{u^2} = 1$. Consequently, the energy of $u$ is fixed as
\begin{equation}
E[u] \defeq \frac{1}{2} \int_{-\pi}^{\pi} u^2 \dx = \pi
\end{equation}

@@

Then the Hamiltonian of \eqref{dimlessKdV} is given by
\begin{equation}
\Ham = 3 \pi \epsup \drat^{-1} \, H_3 - \frac{4 \pi^3}{3 \lamfac^2} \delup^2 \drat \, H_2
\end{equation}
where $\drat$ changes value across the ADC. More specifically, there exists an upstream and a downstream Hamiltonian
\begin{align}
&\Hup = 3 \pi \epsup \, H_3 - \frac{4 \pi^3}{3 \lamfac^2} \delup^2 \, H_2 \\
&\Hdn = 3 \pi \epsup \dratdn^{-1} \, H_3 - \frac{4 \pi^3}{3 \lamfac^2} \delup^2 \dratdn \, H_2
\end{align}
$\Hup$ is a conserved quantity for $t<0$ and $\Hdn$ is a conserved quantity for $t>0$. Each Hamiltonian $\Ham^{\pm}$generates a corresponding Gibbs measure $\Gibbs^{\pm}$. (FILL IN DEFINITIONS)

\vsp{10} 
\subsection{Matching condition}

The statistical matching condition is
\begin{equation}
\meanup{\Hdn} = \meandn{\Hdn}
\end{equation}
An experimental observation is that the incoming skewness is small, thus
\begin{equation}
\meanup{H_3} \approx 0
\end{equation}
which immediately gives
\begin{equation}
\ampp D^{-3/2} \meandn{H_3} = \mu D^{1/2} \left( \meandn{H_2} - \meanup{H_2} \right)
\end{equation}
or equivalently
\begin{equation}
\label{H3H2}
\meandn{H_3} = \frac{\mu D^{2}}{\ampp} \, \Delta \mean{H_2}
\end{equation}
where $\Delta \mean{H_2} =  \meandn{H_2} - \meanup{H_2}$  indicates the difference between the upstream and downstream values.

>>>>

%----------------------------------------------------------------------------%
% Mathematical framework
%----------------------------------------------------------------------------%
\section{The truncated KdV framework}

%----------------------------------------------------------------------------%
% KdV
%----------------------------------------------------------------------------%
\subsection{The Korteweg–de Vries equation with variable depth}
Consider waves propagating unidirectionally in shallow water. Consider the surface displacement $\eta(x,t)$ and the reference frame moving with the local wave speed $\xi = x - ct$, where $c = \sqrt{g \depth}$ is the wave speed, $g$ gravity, and $\depth$ the local depth.
To first-correction in small amplitude, surface displacements are governed by the Korteweg–de Vries equation (KdV), which in dimensional form is given by
\begin{equation}
2 \eta_t + \frac{3 c}{\depth} \eta \eta_{\xi} + \frac{c \depth^2}{3} \eta_{\xi \xi \xi} = 0
\end{equation}
% The coefficients and scales can be verified on Wolfram KdV page (but what about the sign?). I would like an official source as confirmation though.

@@

%----------------------------------------------------------------------------%
% Nondimensionalization
%----------------------------------------------------------------------------%
\subsection{Dimensionless variables}
\label{nondim}

% Table
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{table}[h]%[htbp]
\begin{center}
\caption{Table of parameters}
\label{paramtable}
\begin{tabular}{l l l}
\hline \multicolumn{3} { c }{Parameters that are constant in a single experiment} \\
\hline Description & Notation and definition & Value in experiments \\
\hline
Peak forcing frequency		& $f_p$						& 2 Hz \\
Upstream depth			& $\dup$						& 12.5 cm \\
Downstream depth			& $\ddn$						& 3 cm (and varied) \\
Characteristic wave amplitude	& $a = \mean{\eta^2}^{1/2} $		& 0.03--0.3 cm \\
Upstream wavelength		& $\lamup = \sqrt{g \dup}/f_p$		& 56 cm \\
Amplitude-to-depth ratio		& $\epsup = a / \dup$			& 0.0024--0.024 \\
Depth-to-wavelength ratio		& $\delup = \dup / \lamup$		& 0.22 \\
Depth ratio				& $\dratdn = \ddn/\dup$			& 0.24 (and varied) \\
\hline \multicolumn{3} { c }{Parameters that change value across ADC} \\
\hline Description & Notation and definition \\
\hline
Local depth			& $\depth$			\\
Local wavespeed		& $c = \sqrt{g \depth}$	\\
Local peak wavelength	& $\lam = \sqrt{gd}/f_p$	\\
Local dimensionless depth		& $\drat = \depth/\dup$	
\end{tabular}
\end{center}
\end{table}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

@@

%----------------------------------------------------------------------------%
% Johnson alternative
%----------------------------------------------------------------------------%
\subsubsection{Alternative KdV formulation following Johnson}
Section \ref{nondim} makes the implicit assumption of continuous surface displacement crossing the ADC. This is a very physically motivated assumption, but there is a small amount of debate surrounding the assumption. An alternative is to appeal to Green's law and assume the quantity $\depth^{1/4} \eta$ to be continuous across the ADC, which necessarily implies a small discontinuity in the surface (which of course cannot be true literally), as is the approach taken in Johnson't book. The dimensional KdV equation would then be expressed as
\begin{equation}
2 (\depth^{1/4} \eta)_t + \frac{3 c}{\depth^{5/4}} (\depth^{1/4} \eta) (\depth^{1/4} \eta)_{\xi} + \frac{c \depth^2}{3} (\depth^{1/4} \eta)_{\xi \xi \xi} = 0
\end{equation}
Then, the dimensionless state variable is instead 
\begin{equation}
u = \frac{\depth^{1/4} \, \eta}{ \dup^{1/4} \, \amp}
\end{equation}
and the dimensionless KdV equation takes the form
\begin{equation}
u_t + \frac{3 \pi}{\lamfac} \epsup \drat^{-5/4} \, u u_x + \frac{4 \pi^3}{3 \lamfac^3} \delup^2 \drat \, u_{xxx} = 0
\qquad \text{for } x \in [-\pi,\pi]
\end{equation}
Notice the only difference is a small change in the power of $\drat$ in the second term (changes from $\drat^{-1}$ to $\drat^{-5/4}$).
We will not use this alternative formulation.

%----------------------------------------------------------------------------%
% Hamiltonian
%----------------------------------------------------------------------------%
\subsection{Hamiltonian structure}

@@

We measure the state variable $u$ as the displacement of the surface from equilibrium, so that $\mean{u} = 0$. Consequently, the momentum of $u$ vanishes
\begin{equation}
M[u] \defeq \int_{-\pi}^{\pi} u \dx = 0
\end{equation}
Next, the definition of the characteristic amplitude in \eqref{ampdefn} implies that $\mean{u^2} = 1$. Consequently, the energy of $u$ is fixed as
\begin{equation}
E[u] \defeq \frac{1}{2} \int_{-\pi}^{\pi} u^2 \dx = \pi
\end{equation}
TO DO: Add Fourier stuff including Parseval.
\vsp{4}

@@

Then the Hamiltonian of \eqref{dimlessKdV} is given by
\begin{equation}
\Ham = 3 \pi \epsup \drat^{-1} \, H_3 - \frac{4 \pi^3}{3 \lamfac^2} \delup^2 \drat \, H_2
\end{equation}
where $\drat$ changes value across the ADC. More specifically, there exists an upstream and a downstream Hamiltonian
\begin{align}
&\Hup = 3 \pi \epsup \, H_3 - \frac{4 \pi^3}{3 \lamfac^2} \delup^2 \, H_2 \\
&\Hdn = 3 \pi \epsup \dratdn^{-1} \, H_3 - \frac{4 \pi^3}{3 \lamfac^2} \delup^2 \dratdn \, H_2
\end{align}
$\Hup$ is a conserved quantity for $t<0$ and $\Hdn$ is a conserved quantity for $t>0$. Each Hamiltonian $\Ham^{\pm}$ generates a corresponding Gibbs measure $\Gibbs^{\pm}$ with corresponding ensemble means $\mean{}_{\pm}$. (FILL IN DEFINITIONS)

%----------------------------------------------------------------------------%
% Matching
%----------------------------------------------------------------------------%
\subsection{Matching condition}

The statistical matching condition is
\begin{equation}
\meanup{\Hdn} = \meandn{\Hdn}
\end{equation}
An experimental observation is that the incoming skewness is small, thus
\begin{equation}
\meanup{H_3} \approx 0
\end{equation}
which immediately gives
\begin{equation}
3 \pi \epsup \dratdn^{-1} \meandn{H_3} = \frac{4 \pi^3}{3 \lamfac^2} \delup^2 \dratdn \left( \meandn{H_2} - \meanup{H_2} \right)
\end{equation}
or more simply
\begin{equation}
\label{H3H2}
\meandn{H_3} = \frac{4 \pi^2}{9 \lamfac^2} \frac{\delup^2}{\epsup} \dratdn^2 \,\, \Delta \mean{H_2}
\end{equation}
where $\Delta \mean{H_2} =  \meandn{H_2} - \meanup{H_2}$  indicates the difference between the upstream and downstream values.
To convert to dimensional, i.e.~experimental, values we use
\begin{align}
&\mean{u^3} = \frac{\mean{\eta^3}}{a^3} = 
\frac{\mean{\eta^3}}{\mean{\eta^2}^{3/2}} = \skw(\eta) \\
&\mean{u_x^2} = \left(\frac{\lam}{a} \right)^2 \mean{\eta_x^2} 
= \left(\frac{\lam}{a} \right)^2 \var(\eta_x)
\end{align}
Then \eqref{H3H2} gives the relationship
\begin{equation}
\frac{\skw(\eta)}{\Delta \var(\eta_x)} =
3 \mu D^2 \left( \frac{h_0}{a} \right)^3 = 3 \eps^{-3} \mu D^2
\end{equation}
In particular, the ratio on the left, which can be measured in the experiments, is predicted to scale as the inverse cube of the wave amplitude (as well as the square of the depth ratio).

WORK TO DO HERE WITH LAMBDA

\vsp{10}
OLD VERSION: \\
which immediately gives
\begin{equation}
\ampp D^{-3/2} \meandn{H_3} = \mu D^{1/2} \left( \meandn{H_2} - \meanup{H_2} \right)
\end{equation}
or equivalently
\begin{equation}
\label{H3H2}
\meandn{H_3} = \frac{\mu D^{2}}{\ampp} \, \Delta \mean{H_2}
\end{equation}
where $\Delta \mean{H_2} =  \meandn{H_2} - \meanup{H_2}$  indicates the difference between the upstream and downstream values.
